Computational biology in the 21st century: Algorithms that scale

### 2015-07-28 main.tex
More minor edits

This growth in data poses significant challenges for 
researchers~\cite{marx2013biology}.
Currently, many biological `omics' applications require us to store, access, 
and analyze large libraries of data.
One approach to solving these challenges is to embrace cloud computing.
Google, Inc. and the Broad Institute have collaborated to bring the GATK 
(Genome 
Analysis Toolkit) to the Google cloud (https://cloud.google.com/genomics/gatk).
Amazon Web Services are also commonly used for computational biology 
research and enterprise (e.g., DNANexus)~\cite{schatz2010cloud}.
However, while cloud computing frees researchers from maintaining their own
data centers and provides cost-saving benefits when computing resources are
not needed continuously, it is no panacea.
First and foremost, the computer systems that make up those cloud datacenters
are themselves bound by improvements in semiconductor technology and Moore's
law. Thus, cloud computing does not truly address the problem posed by the
faster-than-Moore's law exponential growth in omics data.
Moreover, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
in West Africa, analysis resources are needed at often-remote field sites.
While it is now possible to bring sequencing equipment and limited computing 
resources to remote sites, internet connectivity is still highly constrained.
Thus, accessing cloud resources for analytics may not be possible.

@@

When a reference sequence is available, NGS reads can be \emph{mapped} onto
this reference (Figure~\ref{fig:ngs-pipeline}).
Mapping allows the differences between the newly-sequenced genome and the 
reference to be analyzed; these differences, or variants, may include single-nucleotide polymorphisms
(SNPs, which are the genetic analogue to bit-flips) (Figure~\ref{fig:snps}), 
insertions or deletions, or larger-scale changes in the genome.
Determining the differences between an individual genome and a reference is
known as \emph{variant calling}.
While reference-based read mapping is a fundamentally simpler problem than
\emph{de novo} assembly, it is still computationally complex, as gigabytes or
terabytes of reads must each be mapped onto the reference genome, which can 
range from millions (for bacteria) to billions (for mammals).
As an example, the ICGC-TCGA Pan Cancer Analysis of Whole Genomes 
(PCAWG)~\cite{weinstein2013cancer} brings together more than 500 top cancer researchers from about 80 institutions in a coordinated manner with the goal of mapping the entire mutational landscape of 37 common cancer types. 
Currently each sample requires seven hours to download. 
Importantly, researchers do not trust the provided mapping, and thus they redo mappings. 
The time spent on mapping is about 50\% of the overall time spent on the sequence analysis pipeline. 
As read mapping is typically the most costly step in NGS analysis pipelines (e.g., GATK), any improvement to existing mappers will immediately accelerate sequence analysis studies on large read datasets.

@@

On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{gillespie2002isolation,forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease~\cite{gevers2014treatment,manichanh2006reduced}, and 
obesity~\cite{greenblum2012metagenomic}.
Metagenomics fundamentally asks what organisms are present, and, in the case
of a microbiome such as the gut, what metabolic functions it can accomplish as
a whole.
One way of addressing this problem is to attempt to map NGS reads from a metagenomic sample
onto a set of reference genomes that are expected to be present.
This is exactly the read-mapping problem discussed above, but with many 
reference genomes, compounding the computational requirements.
A second way is to perform homology search on a protein sequence database;
exact or nearly-exact matches imply the presence of a species, while more 
distant hits may still give clues to function.
For this task, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
The difficulty is that the datasets required to shine any light on these 
questions, namely from ``shotgun'' metagenomics, are gigantic and vastly more 
complex than 16S metagenomic studies that merely classify the bacteria based on 
their divergence with respect to a specific gene.
The massive data results in major identification 
challenges for certain bacterial, as well as viral, species and 
genera~\cite{janda200716s}.

Given a sequenced genome, the next natural questions are what genes (genomic
regions that code for proteins) are present, what structure each resulting
protein takes, and what biological function it performs.
Identifying likely genes is a well-studied problem~\cite{burge1998finding} 
beyond the scope of this Review.
However, determining evolutionary relationships, structure, and function are
at the heart of current research in computational biology.
Since some organisms (known as \emph{model organisms}) are better studied than
others, and evolution is known to conserve sequence, structure, and function,
a powerful approach to determine these attributes is to search for similar
sequences about which more is known. 
This so-called \emph{homology search} entails searching for \emph{approximate}
matches in databases of known gene or protein sequences.
The homology search problem was previously believed to be solved;
Basic Local Alignment Search Tool (BLAST)~\cite{altschul1990basic} had been
the standard tool for performing homology (similarity) search on databases of nucleotide and
protein sequences.
BLAST takes a ``seed-and-extend'' approach; it looks for small, $k$-mer matches
that might lead to longer matches, and greedily extends them, ultimately 
producing a \emph{sequence alignment} between a query and each potential 
database hit.
However, BLAST's running-time scales linearly with the size of the database being
searched, which is problematic as sequence databases continue to grow at a
faster rate than Moore's law.

@@

\section{Acknowledgments}

>>>>

This growth in data poses significant challenges for 
researchers~\cite{marx2013biology}.
Currently, many biological `omics' applications require us to store, access, 
and analyze large libraries of data.
% YWY: `omics' is the only single quote we use. Are we consistent?
One approach to solving these challenges is to embrace cloud computing.
Google, Inc. and the Broad Institute have collaborated to bring the GATK 
(Genome 
Analysis Toolkit) to the Google cloud (https://cloud.google.com/genomics/gatk).
Amazon Web Services are also commonly used for computational biology 
research and enterprise (e.g., DNANexus)~\cite{schatz2010cloud}.
However, while cloud computing frees researchers from maintaining their own
data centers and provides cost-saving benefits when computing resources are
not needed continuously, it is no panacea.
First and foremost, the computer systems that make up those cloud datacenters
are themselves bound by improvements in semiconductor technology and Moore's
law. Thus, cloud computing does not truly address the problem posed by the
faster-than-Moore's law exponential growth in omics data.
Moreover, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
in West Africa, analysis resources are needed at often-remote field sites.
While it is now possible to bring sequencing equipment and limited computing 
resources to remote sites, internet connectivity is still highly constrained.
Thus, accessing cloud resources for analytics may not be possible.

@@

When a reference sequence is available, NGS reads can be \emph{mapped} onto
this reference (Figure~\ref{fig:ngs-pipeline}).
Mapping allows the differences between the newly-sequenced genome and the 
reference to be analyzed; these differences, or variants, may include single-nucleotide polymorphisms
(SNPs, which are the genetic analogue to bit-flips) (Figure~\ref{fig:snps}), 
insertions or deletions, or larger-scale changes in the genome.
Determining the differences between an individual genome and a reference is
known as \emph{variant calling}.
While reference-based read mapping is a fundamentally simpler problem than
\emph{de novo} assembly, it is still computationally complex, as gigabytes or
terabytes of reads must each be mapped onto the reference genome, which can 
range from millions (for bacteria) to billions (for mammals) of base pairs.
As an example, the ICGC-TCGA Pan Cancer Analysis of Whole Genomes 
(PCAWG)~\cite{weinstein2013cancer} brings together more than 500 top cancer researchers from about 80 institutions in a coordinated manner with the goal of mapping the entire mutational landscape of 37 common cancer types. 
Currently each sample requires seven hours to download. 
% YWY: 7 hours for whom? Perhaps "even on an institutional connection"?
Importantly, researchers do not trust the provided mapping, and thus they redo mappings. 
The time spent on mapping is about 50\% of the overall time spent on the sequence analysis pipeline. 
As read mapping is typically the most costly step in NGS analysis pipelines (e.g., GATK), any improvement to existing mappers will immediately accelerate sequence analysis studies on large read datasets.

@@

On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{gillespie2002isolation,forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease~\cite{gevers2014treatment,manichanh2006reduced}, and 
obesity~\cite{greenblum2012metagenomic}.
Metagenomics fundamentally asks what organisms are present, and, in the case
of a microbiome such as the gut, what metabolic functions it can accomplish as
a whole.
One way of addressing this problem is to attempt to map NGS reads from a metagenomic sample
onto a set of reference genomes that are expected to be present.
This is exactly the read-mapping problem discussed above, but with many 
reference genomes, compounding the computational requirements.
A second way is to perform homology search on a protein sequence database;
exact or nearly-exact matches imply the presence of a species, while more 
distant hits may still give clues to function.
For this task, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
% YWY: 16S isn't well introduced
The difficulty is that the datasets required to shine any light on these 
questions, namely from ``shotgun'' metagenomics, are gigantic and vastly more 
complex than the 16S metagenomic studies that merely classify the bacteria based on 
their divergence with respect to a specific gene.
The massive data results in major identification 
challenges for certain bacterial, as well as viral, species and 
genera~\cite{janda200716s}.

Given a sequenced genome, the next natural questions ask what genes (genomic
regions that code for proteins) are present, what structure each resulting
protein takes, and what biological function it performs.
Identifying likely genes is a well-studied problem~\cite{burge1998finding} 
beyond the scope of this Review.
However, determining evolutionary relationships, structure, and function is
at the heart of current research in computational biology.
Since some organisms (known as \emph{model organisms}) are better studied than
others, and evolution is known to conserve sequence, structure, and function,
a powerful approach to determine these attributes is to search for similar
sequences about which more is known. 
This so-called \emph{homology search} entails searching for \emph{approximate}
matches in databases of known gene or protein sequences.
The homology search problem was previously believed to be solved;
Basic Local Alignment Search Tool (BLAST)~\cite{altschul1990basic} had been
the standard tool for performing homology (similarity) search on databases of nucleotide and
protein sequences.
BLAST takes a ``seed-and-extend'' approach; it looks for small, $k$-mer matches
that might lead to longer matches, and greedily extends them, ultimately 
producing a \emph{sequence alignment} between a query and each potential 
database hit.
However, BLAST's running-time scales linearly with the size of the database being
searched, which is problematic as sequence databases continue to grow at a
faster rate than Moore's law.

@@

\section{Acknowledgments}
Y.W.Y. is supported by a Hertz Foundation Fellowship.
